---
title: "Phylogenetic trees with ggtree: Exercises"
output: 
  pdf_document: 
    fig_height: 3
    fig_width: 5
# fontsize: 8pt
# geometry: margin=.35in
---

## Exercise 1

Look at the help again for `?ggtree`, specifically at the `layout=` option. By default, it produces a rectangular layout.

1. Create a slanted phylogenetic tree.
1. Create a circular phylogenetic tree.
1. Create a circular unscaled cladogram with thick red lines.

----

```{r ex1, echo=FALSE, eval=TRUE}
suppressWarnings(suppressPackageStartupMessages(library(ggtree)))
tree <- read.tree(file.path(rprojroot::find_rstudio_root_file(), "data", "tree_newick.nwk"))
ggtree(tree, layout="slanted")
```

```{r ex1b, echo=FALSE, eval=TRUE}
ggtree(tree, layout="circular")
ggtree(tree, layout="circular", branch.length="none", color="red", size=3)
```

\newpage

## Exercise 2

Similar to how we change the aesthetics for the tree inside the `ggtree()` call, we can also change the aesthetics of the points themselves by passing graphical parameters inside the `geom_nodepoint()` or `geom_tippoint()` calls. Create a phylogeny with the following aesthetic characteristics:

- tips labeled in purple
- purple-colored diamond-shape tip points (hint: Google search "R point characters")
- large semitransparent yellow node points (hint: `alpha=`)
- Add a title with `+ ggtitle(...)`

----

```{r ex2, echo=FALSE, warning=FALSE, message=FALSE}
suppressWarnings(suppressPackageStartupMessages(library(ggtree)))
tree <- read.tree(file.path(rprojroot::find_rstudio_root_file(), "data", "tree_newick.nwk"))
p <- ggtree(tree) 
p + 
  geom_tiplab(color="darkorchid", size=5) + 
  geom_tippoint(color="darkorchid", size=2, shape=18) + 
  geom_nodepoint(color="goldenrod", size=4, alpha=1/2) + 
  ggtitle("Exercise 2 Figure: Not the prettiest phylogenetic aesthetics, but it'll do.")
```

\newpage

## Exercise 3

Produce the figure below.

1. First, find what the MRCA is for taxa **B+C**, and taxa **L+J**. You can do this in one of two ways:
    a. Easiest: use `MRCA(tree, tip=c("taxon1", "taxon2"))` for B/C and L/J separately.
    b. Alternatively: use `ggtree(tree) + geom_text(aes(label=node), hjust=-.3)` to see what the node labels are on the plot. You might also add tip labels here too. 
1. Draw the tree with `ggtree(tree)`.
1. Add tip labels.
1. Highlight these clades with separate colors.
1. Add a clade label to the larger superclade (node=17) that we saw before that includes A, B, C, D, and E. You'll probably need an offset to get this looking right.
1. Link taxa C to E, and G to J with a dashed gray line (hint: get the geom working first, then try changing the aesthetics. You'll need `linetype=2` somewhere in the `geom_taxalink()`).
1. Add a scale bar to the bottom by changing the theme.
1. Add a title.
1. Optionally, go back to the original `ggtree(tree, ...)` call and change the layout to `"circular"`.

----

```{r ex3, echo=FALSE, results="hide", warning=FALSE, message=FALSE, fig.width=8, fig.height=6.5}
suppressWarnings(suppressPackageStartupMessages(library(ggtree)))
tree <- read.tree(file.path(rprojroot::find_rstudio_root_file(), "data", "tree_newick.nwk"))
MRCA(tree, c("B", "C"))
MRCA(tree, c("L", "J"))
ggtree(tree) + 
  geom_tiplab() + 
  geom_hilight(node=19, fill="blue3") + 
  geom_hilight(node=23, fill="orange2") + 
  geom_cladelabel(node=17, label="Superclade 17", color="red3", offset=.8) + 
  geom_taxalink("C", "E", color="gray50", linetype=2) + 
  geom_taxalink("G", "J", color="gray50", linetype=2) + 
  theme_tree2(fgcolor="gray70") + 
  ggtitle("Exercise 3 title: Not sure what we're trying to show here...")
  
```
